Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2018-02-02 paper.tex
Sweep TODO notes and assign remaining ones to coauthors

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries, so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{ruff1996-leakage_part1,ross2003-leakage_part2,
  deat2013-leakage, pere2016-asl, radi2016-dynamical}
\footnote{TODO: cite others}
capture some of the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\footnote{TODO: discuss mc more clearly}
Monte Carlo methods like those
used in supernova simulations \cite{abdi2012-monte_carlo}
and stationary models of accretion disks \cite{rich2015-monte_carlo}
are an excellent tool but require large computational resources
because of their global treatment of the problem.

@@

The ray tracing framework
is conceptually simple because it solves the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann}) along characteristics,
reducing it to a one-dimensional ordinary differential equation.
It is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
with the history of that light-cone truncated at large optical depth.
\footnote{TODO: state scaling $\mathscr{O}(n)$?}
It easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.
And it parallelizes trivially by ray, since each ray is computed
independently.

@@

%\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
\footnote{TODO: cite}
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions,
for example neutrino oscillations and neutrino-antineutrino annihilation,
\footnote{TODO: cite}
depend sensitively on the angular distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\footnote{TODO: justify these statements}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

@@

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\footnote{TODO: clarify how momenta are decoupled}
When solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$,
we have no information about the distribution at different momenta
$f^{\nu_\sigma}(x^\alpha;p'_\beta)$,
or about the distribution of the relevant antineutrino species
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$;
these missing data are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

@@

In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present an application
in the dynamical environment following the merger of two neutron stars,
computing neutrino flavor oscillation along an outgoing ray,
including the effects of coherent forward scattering with ambient neutrinos.
In Sec.~\ref{sec:conclusions} we summarize our work and anticipate improvements.

@@

In Minkowski spacetime, and for a stationary observer
$u_\beta\rightarrow(-1,0,0,0)$, we would have $s=0$, $q=1$,
and $\varepsilon=-p_t$. In that case $\cos A$ may be identified
\footnote{TODO: cite someone that uses $\mu$}
with the familiar forward direction cosine $\mu$.

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\footnote{TODO: point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x} below)
and $C[f]$ is the source term arising from interactions with the medium.
By Eqn.~\ref{eqn:geodesic_x},
the affine parameter has dimension ${\rm length}\,{\rm energy}^{-1}$,
so the source term has dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
for the reactions
\begin{align}
  \nu_e + n            &\leftrightarrow e^- + p,\nonumber\\
  \bar{\nu}_e + p      &\leftrightarrow e^+ + n,\nonumber\\
  \nu_e + {}^AZ        &\leftrightarrow e^- + {}^A(Z+1),\nonumber\\
  \bar{\nu}_e + {}^AZ  &\leftrightarrow e^+ + {}^A(Z-1),\nonumber
\end{align}
with ${}^AZ$ representing a nucleus of mass number $A$ and charge $Z$.
In terms of the emissivity $j(\varepsilon)$ describing number of neutrinos
of energy $\varepsilon$ emitted per length,
and the absorption opacity $\chi_a(\varepsilon)$
describing the number absorbed per length,
the coefficients are
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the distribution function of neutrinos in radiative
equilibrium with the matter, i.e. the Fermi-Dirac distribution function
\begin{equation}
  \label{eqn:feq}
  f^{\rm eq}(\varepsilon) \equiv
  \left(1+e^{\varepsilon/(k_{\rm B} T) -\eta_\nu}\right)^{-1},
\end{equation}
with the neutrino chemical potentials dependent on the local density,
temperature, and composition of the fluid via the neutron, proton,
and electron chemical potentials:
$\eta_{\nu_e}=-\eta_{\bar{\nu}_e}=\eta_p-\eta_n+\eta_{e^-}$ and
$\eta_{\nu_x}=0$.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption \cite{burr2006-neutrino_opacities}.
By detailed balance of the absorption/emission reactions, we may alternatively
write the emissivity in terms of the equilibrium distribution function:
\begin{equation}
  \label{eqn:ae_emissivity_feq}
  \mathscr{E}_{\rm AE}(\varepsilon)
  = \mathscr{K}_{\rm AE}(\varepsilon) f^{\rm eq}(\varepsilon).
\end{equation}
See App.~\ref{ssec:sources_ae} for details.
\footnote{TODO: identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \ell'^\gamma,
\end{align}
defining the specific energy density,
specific momentum density, and
specific radiation pressure tensor, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\footnote{TODO: relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos a)\,db$.
We also make use of the specific number density and specific number flux defined
\begin{align}
  \label{eqn:G}
  G(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta), \\
  \label{eqn:Ka}
  K^\mu(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu.
\end{align}

@@

We integrate Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} with adaptive step sizes,
using the 3rd order Runge-Kutta algorithm
which produces an error estimate by comparing the 3rd and 2nd order solutions.
After each step is taken, the errors for each of the nine variables of
Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} are compared.
If the maximum of these errors exceeds some absolute or relative threshold
\footnote{TODO: specify ScaledAbsRel algorithm}
the step size is decreased and the step recomputed,
if it is below some threshold the next step size is increased.
% see OdeIntegrators::AdaptiveDense
% and OdeSteppers::RungeKutta3
% and OdeSteppers::DormandPrince5
% and OdeController
% and OdeControllers::ProportionalIntegral_Base::StepSizeToAttempt
% and OdeErrorMeasures::ScaledAbsRel
\footnote{TODO: specify}

@@

To test our algorithm, we integrate the ray tracing equations
(Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau}) for various observers
in the following configurations. This suite of configurations
may be understood as a hierarchy of increasing physical realism:
beginning with a homogeneous medium of effectively
infinite extent (i.e.\ optically thick) and progressing to a model of a
1D pre-supernova, post-bounce collapse profile
evolved using an M1 transport hydrodynamical simulation.
In the pre-supernova model, we compare ray tracing distributions
to those calculated in a Monte Carlo transport simulation.
Finally, we calculate global measures of the neutrino fields outside of the
hypermassive neutron star and disk formed in a binary neutron star
merger simulated by \cite{fouc2016-m1_nsns, fouc2016-m1_evolve_n},
and compare ray tracing results to those from the M1 transport simulation.

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE,scat}+f_{\rm SE,scat}$
(labeled \emph{scat} since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE,noscat}$
(labeled \emph{noscat} since we ignore elastic scattering).
There is no simple relation between these forms
(i.e.\ $f_{\rm AE,scat} \neq f_{\rm AE,noscat}$)
because $f_{\rm AE}$ is integrated using the absorption plus scattering optical depth
in the first form and solely the absorption optical depth in the second form.
The differences between the two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the \emph{noscat} form of the equations has been the standard for
\footnote{TODO: explain why; cite others}
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter in a Minkowski spacetime
representative of the fluid thermodynamic state at a radius of 50~km
in the test presented in Sec.~\ref{ssec:test_collapse}---a
massive collapsed star following core bounce---where
$\varrho=10^{11}\,{\rm g}\,{\rm cm}^{-3}$, $T=3.7$~MeV, and $Y_e=0.12$.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).
The opacity table is computed using \lstinline{NuLib} from cross-sections
neglecting weak magnetism
\footnote{TODO: anything else?}
and is stored on a grid covering energy, density, temperature, and composition
ranges spanning $\varepsilon_i\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,6.3\times10^{15}]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,200]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.
Interpolation is performed linearly.

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}, at the single energy
$\varepsilon=11.25$~MeV. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t \frac{dt'}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\footnote{TODO: could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

@@

As Fig.~\ref{fig:mfps-50km} reveals, integrating the distribution functions
over the energies $\varepsilon\in[0,100]$~MeV
probes the numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\footnote{TODO: discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

We sample the distribution function $f(\tilde{\varepsilon},\cos A,B)$
with ray tracing
over a uniform grid of 40 points in energy $\tilde{\varepsilon}\in[0,100]$
and 30 points in angle $\cos A\in[-1,1]$, holding fixed $B=\pi$.
Results are shown in Fig.~\ref{fig:avg_eps_doppler} for
the velocities $v=\{0,0.1,0.8\}$.
The ray tracing results are computed from total densities in each angular bin,
that is
\begin{equation}
  \label{eqn:avg_eps_per_cosA}
  \langle \tilde{\varepsilon} \rangle (\cos A) =
  \frac{\tilde{J}(\cos A)}{\tilde{G}(\cos A)},
\end{equation}
with the Eulerian energy density per angular bin given by
\begin{equation}
  \label{eqn:J_per_cosA}
  \tilde{J}(\cos A) =
  \frac{1}{(2\pi)^3} \int_0^{\infty} d\tilde{\varepsilon} \, \tilde{\varepsilon}^3
  \int_0^{2\pi} dB \, f(\tilde{\varepsilon},\cos A,B),
\end{equation}
and $\tilde{G}(\cos A)$ defined analogously.
\footnote{TODO: add metric terms for these integrals}
Results are identical for all \emph{scat} and \emph{noscat} methods.
In Fig.~\ref{fig:avg_eps_doppler}
we see the common features of a red-shifted spectrum for receding fluid
($\cos A\ll0$),
a blue-shifted spectrum for approaching fluid
($\cos A\gg0$),
and a slightly red-shifted spectrum for fluid moving transverse to the
observer ($\cos A=0$).

@@

\subsection{Idealized Compact Star:
  Testing Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Unfinished: describe and point readers to Asano-Fukuyama test presented
in thesis ~\cite[Sec.~4.3.3]{deat2015-thesis}.
Also compare a few geodesics in Kerr to analytic.
\footnote{TODO: compute some geodesics in Kerr}

>>>>

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries, so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{ruff1996-leakage_part1,ross2003-leakage_part2,
  deat2013-leakage, pere2016-asl, radi2016-dynamical}
\footnote{TODO (Evan): cite others}
capture some of the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
Monte Carlo methods like those
used in supernova simulations \cite{abdi2012-monte_carlo}
and stationary models of accretion disks \cite{rich2015-monte_carlo}
are an excellent tool but require large computational resources
because of their global treatment of the problem.
\footnote{TODO (Evan): discuss mc more clearly}

@@

The ray tracing framework
is conceptually simple because it solves the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann}) along characteristics,
reducing it to a one-dimensional ordinary differential equation.
It is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
with the history of that light-cone truncated at large optical depth.
It easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.
And it parallelizes trivially by ray, since each ray is computed
independently.

@@

%\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions,
for example neutrino oscillations \footnote{TODO (Gail): cite an example study}
and neutrino-antineutrino annihilation \cite{asan2000-nunubar},
depend sensitively on the angular distribution.
And spectral changes can strongly affect the nucleosynthesis.
occuring in the ejected and irradiated material
\footnote{TODO (Evan, Francois, Gail): cite energy-dependent
neutrino+nucleosynthesis study}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

@@

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
When solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$,
we have no information about the distribution at different momenta
$f^{\nu_\sigma}(x^\alpha;p'_\beta)$,
or about the distribution of the relevant antineutrino species
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$;
these missing data are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

@@

In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present neutrino fields in
the dynamical environment following the merger of two neutron stars,
and compute neutrino flavor oscillation along an outgoing ray,
including the effects of coherent forward scattering with ambient neutrinos.
In Sec.~\ref{sec:conclusions} we summarize our work and anticipate improvements.

@@

In Minkowski spacetime, and for a stationary observer
$u_\beta\rightarrow(-1,0,0,0)$, we would have $s=0$, $q=1$,
and $\varepsilon=-p_t$. In that case $\cos A$ may be identified
with the familiar forward direction cosine $\mu$
\citep{shu1991-book_radiation,miha1999-foundations}.

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
In the limit of large oscillation lengths (see Sec.~\ref{ssec:V_nunu}),
neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x} below)
and $C[f]$ is the source term arising from interactions with the medium.
As we will show below in Eqn.~\ref{eqn:proper_length},
the affine parameter has dimension ${\rm length}\,{\rm energy}^{-1}$,
so the source term has dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
for the reactions
\begin{align}
  \nu_e + n            &\leftrightarrow e^- + p,\nonumber\\
  \bar{\nu}_e + p      &\leftrightarrow e^+ + n,\nonumber\\
  \nu_e + {}^AZ        &\leftrightarrow e^- + {}^A(Z+1),\nonumber\\
  \bar{\nu}_e + {}^AZ  &\leftrightarrow e^+ + {}^A(Z-1),\nonumber
\end{align}
with ${}^AZ$ representing a nucleus of mass number $A$ and charge $Z$.
In terms of the emissivity $j(\varepsilon)$ describing number of neutrinos
of energy $\varepsilon$ emitted per length,
and the absorption opacity $\chi_a(\varepsilon)$
describing the number absorbed per length,
the coefficients are
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the distribution function of neutrinos in radiative
equilibrium with the matter, i.e. the Fermi-Dirac distribution function
\begin{equation}
  \label{eqn:feq}
  f^{\rm eq}(\varepsilon) \equiv
  \left(1+e^{\varepsilon/(k_{\rm B} T) -\eta_\nu}\right)^{-1},
\end{equation}
with the neutrino chemical potentials dependent on the local density,
temperature, and composition of the fluid via the neutron, proton,
and electron chemical potentials:
$\eta_{\nu_e}=-\eta_{\bar{\nu}_e}=\eta_p-\eta_n+\eta_{e^-}$ and
$\eta_{\nu_x}=0$.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption \cite{burr2006-neutrino_opacities}.
By detailed balance of the absorption/emission reactions, we may alternatively
write the emissivity in terms of the equilibrium distribution function:
\begin{equation}
  \label{eqn:ae_emissivity_feq}
  \mathscr{E}_{\rm AE}(\varepsilon)
  = \mathscr{K}_{\rm AE}(\varepsilon) f^{\rm eq}(\varepsilon).
\end{equation}
Note that the stimulated absorption coefficient $\mathscr{K}_{\rm AE}$ is
identical to the coefficient $\kappa^*$
defined in \cite{hari2010-gr_nunubar_collapsar}.
See App.~\ref{ssec:sources_ae} for details.

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \frac{\varepsilon^3}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu \ell'^\gamma,
\end{align}
defining the specific energy density,
specific momentum density, and
specific radiation pressure tensor, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos a)\,db$.
We also make use of the specific number density and specific number flux defined
\begin{align}
  \label{eqn:G}
  G(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta), \\
  \label{eqn:Ka}
  K^\mu(\varepsilon) &=
  \frac{\varepsilon^2}{(2\pi)^3} \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell'^\mu.
\end{align}

@@

We integrate Eqns.~\ref{eqn:num_x}--\ref{eqn:num_fse} with adaptive step sizes,
using the 3rd order Runge-Kutta algorithm
which produces an error estimate by comparing the 3rd and 2nd order solutions.
After each step is taken, the errors for each of the nine variables of
Eqns.~\ref{eqn:num_x}--\ref{eqn:num_fse} are compared to an absolute and
a relative threshold.
If the error in any variable exceeds its threshold
the step size is decreased and the step recomputed;
if all errors are below threshold, the next step size is increased.
% see OdeIntegrators::AdaptiveDense
% and OdeSteppers::RungeKutta3
% and OdeSteppers::DormandPrince5
% and OdeController
% and OdeControllers::ProportionalIntegral_Base::StepSizeToAttempt
% and OdeErrorMeasures::ScaledAbsRel
% For example, the parameters I use in the nsns data:
%OdeErrorMeasure = ScaledAbsRel
%     (
%      ATol = 0.000555555555555556;
%      RTol = 0.000555555555555556;
%      ScaledVars = q,p,s0,sn,tau,fa,fs;
%      Scales = 4,1,4,4,1e-20,1e-20,1e-20;
In practice, the controlling errors come from the radiation variables
$\tau$, $f_{\rm AE}$, and $f_{\rm SE}$,
for which the relative tolerances are set to $6\times10^{-4}$,
and the absolute tolerances are set to $6\times10^{-24}$.

@@

To test the algorithm, we integrate the ray tracing equations
(Eqns.~\ref{eqn:num_x}--\ref{eqn:num_fse}) for various observers
in the following configurations. This suite of configurations
defines a hierarchy of increasing physical realism:
beginning with a homogeneous medium of effectively
infinite extent (i.e.\ optically thick) and progressing to a model of a
1D pre-supernova, post-bounce collapse profile
evolved using an M1 transport hydrodynamical simulation.
In the pre-supernova model, we compare ray tracing distributions
to those calculated in a Monte Carlo transport simulation.

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
\begin{itemize}
\item the \emph{scat} form including elastic scattering,
  for which the solution is
  $f=f_{\rm AE,scat}+f_{\rm SE,scat}$,
\item the \emph{noscat} form treating only absorption/emission interactions
  by setting $\mathscr{K}_{\rm SE}=0$,
  for which the solution is
  $f=f_{\rm AE,noscat}$.
\end{itemize}
Note that $f_{\rm AE,scat} \neq f_{\rm AE,noscat}$ because $f_{\rm AE}$
is integrated using the absorption plus scattering optical depth in
the first form and using only the absorption optical depth in the second form.
So there is no simple algebraic relation between the \emph{scat} and
\emph{noscat} distribution functions.
The differences between the two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous} below.
Before this work, the \emph{noscat} form of the equations was the standard
for ray tracing, though many authors included the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter in a Minkowski spacetime
representative of the fluid thermodynamic state at a radius of 50~km
in the test presented in Sec.~\ref{ssec:test_collapse}---a
massive collapsed star following core bounce---where
$\varrho=10^{11}\,{\rm g}\,{\rm cm}^{-3}$, $T=3.7$~MeV, and $Y_e=0.12$.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).
The opacity table is computed using \lstinline{NuLib} from cross-sections
neglecting weak magnetism
\footnote{TODO (Evan): neglecting anything else?}
and is stored on a grid covering energy, density, temperature, and composition
ranges spanning $\varepsilon_i\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,6.3\times10^{15}]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,200]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.
Interpolation is performed linearly.

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}, at the single energy
$\varepsilon=11.25$~MeV. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t \frac{dt'}{\varepsilon}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

@@

As Fig.~\ref{fig:mfps-50km} reveals, integrating the distribution functions
over the energies $\varepsilon\in[0,100]$~MeV
probes the numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\footnote{TODO (Matt, Brett): look at Brett's tests and make clearer statement}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

We sample the distribution function $f(\tilde{\varepsilon},\cos A,B)$
with ray tracing
over a uniform grid of 40 points in energy $\tilde{\varepsilon}\in[0,100]$
and 30 points in angle $\cos A\in[-1,1]$, holding fixed $B=\pi$.
Results are shown in Fig.~\ref{fig:avg_eps_doppler} for
the velocities $v=\{0,0.1,0.8\}$.
The ray tracing results are computed from total densities in each angular bin,
that is
\begin{equation}
  \label{eqn:avg_eps_per_cosA}
  \langle \tilde{\varepsilon} \rangle (\cos A) =
  \frac{\tilde{J}(\cos A)}{\tilde{G}(\cos A)},
\end{equation}
with the Eulerian energy density per angular bin given by
\begin{equation}
  \label{eqn:J_per_cosA}
  \tilde{J}(\cos A) =
  \frac{1}{(2\pi)^3} \int_0^{\infty} d\tilde{\varepsilon} \, \tilde{\varepsilon}^3
  \int_0^{2\pi} dB \, f(\tilde{\varepsilon},\cos A,B),
\end{equation}
and $\tilde{G}(\cos A)$ defined analogously.
Results are identical for all \emph{scat} and \emph{noscat} methods.
In Fig.~\ref{fig:avg_eps_doppler}
we see the common features of a red-shifted spectrum for receding fluid
($\cos A\ll0$),
a blue-shifted spectrum for approaching fluid
($\cos A\gg0$),
and a slightly red-shifted spectrum for fluid moving transverse to the
observer ($\cos A=0$).

@@

\subsection{Idealized Compact Star:
  Testing Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Unfinished: describe and point readers to Asano-Fukuyama test presented
in thesis ~\cite[Sec.~4.3.3]{deat2015-thesis}.
Also compare a few geodesics in Kerr to analytic.
\footnote{TODO (Brett): compute some geodesics in Kerr}

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
To test our scattering treatments we calculate neutrino fields outside a
collapsed 15~${\rm M}_\odot$ star, 100~ms after core bounce,
comparing ray tracing fields to those from a Monte Carlo transport calculation.
Elastic scattering interior to the shock at $r\approx150$~km
significantly modifies the neutrinos' spectra,
and the extended envelope outside the shock becomes a source of
higher-than-average-energy neutrinos.
